Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-01-13 manuscript/manuscript.Rmd
WIP on manuscript.

---
title: "Shift-sensitive Power Transformations for Transforming Data to Normality"
author: "Alex Zwanenburg"
date: "2023-01-06"
output: 
  pdf_document:
    latex_engine: lualatex
bibliography: "refs.bib"
header-includes:
   - \usepackage{amsmath}
---

@@

Power transformation does not guarantee that transformed variables are normally distributed. In fact, depending on location and scale of the variable, power transformation may reduce normality, as shown in Figure.

``` {r reduced-normality, echo=FALSE, fig.cap=cap}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values are randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and even breaks down due to numerical issues for $\\mu > 10^3$.")

@@

@@

Figure showing Ames dataset feature "Year_Built". Panels: Histogram of feature; Q-Q plot of original feature, normal Box-Cox, robust Box-Cox, normal Yeo-Johnson, robust Yeo-Johnson.

We make the following contributions:

@@

@@

>>>>

---
title: "Shift-sensitive Power Transformations for Transforming Data to Normality"
author: "Alex Zwanenburg"
date: "2023-01-06"
output: 
  pdf_document:
    latex_engine: lualatex
bibliography: "refs.bib"
header-includes:
   - \usepackage{amsmath}
   - \DeclareMathOperator*{\argmax}{argmax}
   - \DeclareMathOperator{\sgn}{sgn}
---

@@

Power transformation does not guarantee that transformed variables are normally distributed. In fact, depending on location and scale of the variable, power transformation may reduce normality, as shown in Figure **INSERT REFERENCE**. This is problematic for automated power transformations, for three reasons. First, normality is often desired in machine learning applications. Second, a large negative or positve $\lambda$-parameter may lead to numeric issues. Third, we currently do not have a test to automatically reject poor transformations. Statistical tests for normality exist, such as the Shapiro-Wilk test [@Shapiro1965-zd]. Given sufficiently large sample sizes, these tests can detect trivial deviations from normality and are therefore not well-suited for practical applications of automated power transformations.

```{r reduced-normality, echo=FALSE, fig.cap=cap}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values are randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and even breaks down due to numerical issues for $\\mu > 10^3$.")

@@

# Non-standard evaluation
mu <- lambda <- method <- NULL

@@

Here we make the following contributions:

@@

## Shift-sensitive power transformation

Box-Cox and Yeo-Johnson transformations are modified by adding a shift parameter $x_0$. The modified Box-Cox transformation of a feature $x$ under transformation parameter $\lambda$ and shift parameter $x_0$ is then:

```{=latex}
\begin{equation}
\phi_{\text{BC}}^{\lambda, x_0} (x) = 
\begin{cases}
\left( \left(x - x_0 \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x - x_0) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
Where $x - x_0 > 0$. Likewise, the modified Yeo-Johnson transformation of a feature $x$ under transformation parameter $\lambda$ and shift parameter $x_0$ is:

```{=latex}
\begin{equation}
\phi_{\text{YJ}}^{\lambda, x_0} (x) = 
\begin{cases}
\left( \left( 1 + x - x_0\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x - x_0 \geq 0\\
\log(1 + x - x_0) & \text{if } \lambda = 0 \text{ and } x - x_0 \geq 0\\
-\left( \left( 1 - x - x_0\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x - x_0< 0\\
-\log(1 - x - x_0) & \text{if } \lambda = 2 \text{ and } x - x_0 < 0
\end{cases}
\end{equation}
```
For both modified transformations, $\lambda$ and $x_0$ are obtained by maximising the normal log-likelihood function. For the modified Box-Cox transformation, and ignoring constant terms, this can be written as:

```{=latex}
\begin{equation}
\left\{ \hat{\lambda}^{\text{BC}}, \hat{x}_0^{\text{BC}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \log(x_i - x_0)
\end{equation}
```
where $\sigma^2$ is the variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (x)$. Similarly for the modified Yeo-Johnson transformation:

```{=latex}
\begin{equation}
\left\{ \hat{\lambda}^{\text{YJ}}, \hat{x}_0^{\text{YJ}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \sgn(x_i - x_0) \log(1 + |x_i - x_0|)
\end{equation}
```
where $\sigma^2$ is the variance of Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0} (x)$.

## Robust shift-sensitive power transformations

Real-world data may contain outliers, to which maximum likelihood estimation can be sensitive. Their presence may lead to poor transformations to normality, as shown in Figure **INSERT REFERENCE**. As indicated by @Raymaekers2021-kq, the general aim of power transformations should be to transform non-outlier data to normality, i.e. achieve *central normality*. To achieve this, they devised an iterative procedure to find a robust estimate of the transformation parameter $\lambda$. Because $\lambda$ is updated during this procedure, simultaneous estimation of $\lambda$ and $x_0$ is not possible.

@@

In their work, Box and Cox already mention a transformation with a shift parameter, but preferred the now well-known version in Eq. **INSERT REFERENCE** for the theoretical analysis in their paper [@Box1964-mz].
